---
title: "When Your Predictor Disappears: Mediation vs. Multiple Regression"
author: "Rahul Venugopal"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: lumen
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 300
)
```

---

## The Puzzle

Bill length predicts body mass (r = 0.60). That's clear and significant. But add flipper
length to the model and the bill-length coefficient collapses — from **β = 0.595** to a
non-significant **β = 0.041**. Did bill length suddenly stop mattering? No. Something
subtler is happening: bill length's effect on body mass is being **carried** by flipper
length, not blocked by it.

This post walks through that story in two complementary ways — **multiple regression** to
quantify the coefficient collapse, and **mediation analysis** to formally decompose *why*
it happens.

---

## Setup

```{r libraries}
library(palmerpenguins)
library(mediation)
library(tidyverse)
library(broom)
library(patchwork)
```

```{r data}
# Drop rows with NAs in our three variables, then standardise
penguins_clean <- na.omit(penguins[, c("body_mass_g", "flipper_length_mm", "bill_length_mm")])
penguins_s     <- as.data.frame(scale(penguins_clean))
```

Standardising (z-scoring) means every coefficient below is in **standard-deviation units**,
so comparisons across models are apples-to-apples.

---

## Figure 1 — All three variables are correlated

Before modelling anything, it is worth seeing the raw relationships. All three pairwise
correlations are positive and substantial.

```{r associations, fig.width=12, fig.height=5}
theme_blog <- theme_minimal(base_size = 15) +
  theme(
    plot.title       = element_text(face = "bold", size = 17, margin = margin(b = 6)),
    plot.subtitle    = element_text(color = "grey40", size = 11, margin = margin(b = 10)),
    plot.caption     = element_text(color = "grey50", size = 9,  margin = margin(t = 10)),
    axis.title       = element_text(size = 14),
    panel.grid.minor = element_blank(),
    plot.background  = element_rect(fill = "white", color = NA)
  )

scatter_bm_bl <- ggplot(penguins_s, aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point(alpha = 0.4, color = "#2171b5", size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, color = "#08519c", linewidth = 0.9) +
  labs(x = "Bill length (SD)", y = "Body mass (SD)",
       title = "Bill length vs Body mass",
       subtitle = paste0("r = ", round(cor(penguins_s$bill_length_mm, penguins_s$body_mass_g), 2))) +
  theme_blog

scatter_fl_bm <- ggplot(penguins_s, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(alpha = 0.4, color = "#238b45", size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, color = "#006d2c", linewidth = 0.9) +
  labs(x = "Flipper length (SD)", y = "Body mass (SD)",
       title = "Flipper length vs Body mass",
       subtitle = paste0("r = ", round(cor(penguins_s$flipper_length_mm, penguins_s$body_mass_g), 2))) +
  theme_blog

scatter_bl_fl <- ggplot(penguins_s, aes(x = bill_length_mm, y = flipper_length_mm)) +
  geom_point(alpha = 0.4, color = "#d94801", size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, color = "#a63603", linewidth = 0.9) +
  labs(x = "Bill length (SD)", y = "Flipper length (SD)",
       title = "Bill length vs Flipper length",
       subtitle = paste0("r = ", round(cor(penguins_s$bill_length_mm, penguins_s$flipper_length_mm), 2))) +
  theme_blog

scatter_bm_bl + scatter_fl_bm + scatter_bl_fl +
  plot_annotation(
    title    = "All three variables are correlated",
    subtitle = "The puzzle: if bill and flipper are both correlated with body mass, what happens when we put them together?",
    caption  = "Data: palmerpenguins | Standardised variables",
    theme    = theme(
      plot.title      = element_text(face = "bold", size = 14),
      plot.subtitle   = element_text(color = "grey40", size = 11),
      plot.background = element_rect(fill = "white", color = NA)
    )
  )
```

The key tension: bill length (r = 0.60) and flipper length (r = 0.87) are both associated
with body mass, *and* they are correlated with each other (r = 0.66). When correlated
predictors compete in the same model, the coefficients can shift dramatically.

---

## Figure 2 — The coefficient collapse

Fitting two models isolates the effect:

```{r models}
model_simple   <- lm(body_mass_g ~ bill_length_mm,                    data = penguins_s)
model_multiple <- lm(body_mass_g ~ bill_length_mm + flipper_length_mm, data = penguins_s)

summary(model_simple)
summary(model_multiple)
```

```{r coefficient-collapse, fig.width=9, fig.height=7}
coef_df <- bind_rows(
  tidy(model_simple,   conf.int = TRUE) %>% mutate(model = "Simple: body_mass ~\nbill_length"),
  tidy(model_multiple, conf.int = TRUE) %>% mutate(model = "Multiple: body_mass ~\nbill_length + flipper_length")
) %>%
  filter(term == "bill_length_mm") %>%
  mutate(
    significant = p.value < 0.05,
    model = factor(model, levels = c(
      "Simple: body_mass ~\nbill_length",
      "Multiple: body_mass ~\nbill_length + flipper_length"
    )),
    label = paste0("β = ", round(estimate, 3),
                   "\np ", ifelse(significant, "< 0.05", "= 0.332"))
  )

ggplot(coef_df, aes(x = model, y = estimate, color = significant)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60", linewidth = 0.7) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.12, linewidth = 0.9) +
  geom_point(size = 5) +
  geom_text(aes(label = label), vjust = -2.25, size = 3.8, fontface = "bold", show.legend = FALSE) +
  scale_color_manual(
    values = c("TRUE" = "#08519c", "FALSE" = "#cb181d"),
    labels = c("TRUE" = "Significant", "FALSE" = "Not significant"),
    name   = NULL
  ) +
  scale_y_continuous(limits = c(-0.15, 0.85)) +
  labs(
    x       = NULL,
    y       = "Coefficient for bill_length_mm (95% CI)",
    title   = "Figure 2 — The coefficient collapse",
    subtitle = "Bill length's effect on body mass drops from 0.595 to 0.041 once flipper length enters the model",
    caption  = "Data: palmerpenguins | Standardised variables"
  ) +
  theme_blog +
  theme(legend.position = "bottom")
```

Regression can show *that* the coefficient collapsed. It cannot, on its own, tell us *why*.
That requires thinking about causal structure — which is where mediation comes in.

---

## Mediation analysis — decomposing the effect

Baron & Kenny's framework decomposes the total effect of X → Y into:

| Path | Description |
|------|-------------|
| **c**  | Total effect of X on Y (no mediator) |
| **a**  | Effect of X on mediator M |
| **b**  | Effect of M on Y, controlling for X |
| **c′** | Direct effect of X on Y, controlling for M |
| **a × b** | Indirect (mediated) effect |

Here: **X** = bill length · **M** = flipper length · **Y** = body mass.

```{r mediation-models}
model_c  <- lm(body_mass_g       ~ bill_length_mm,                    data = penguins_s)  # total
model_a  <- lm(flipper_length_mm ~ bill_length_mm,                    data = penguins_s)  # path a
model_bc <- lm(body_mass_g       ~ bill_length_mm + flipper_length_mm, data = penguins_s)  # paths b & c'
```

```{r manual-decomposition}
a   <- coef(model_a)["bill_length_mm"]
b   <- coef(model_bc)["flipper_length_mm"]
c   <- coef(model_c)["bill_length_mm"]
c_p <- coef(model_bc)["bill_length_mm"]

cat(sprintf(
  "  path c   (total effect, bill → body_mass):              %.3f\n
  path a   (bill_length → flipper_length):                 %.3f\n
  path b   (flipper_length → body_mass | bill_length):     %.3f\n
  path c'  (direct: bill_length → body_mass | flipper):    %.3f\n
  indirect effect a×b (mediated via flipper_length):       %.3f\n
  proportion mediated (a×b / c):                           %.1f%%\n",
  c, a, b, c_p, a * b, 100 * (a * b) / c
))
```

93% of bill length's association with body mass is carried through flipper length. The
direct path c′ is near-zero and non-significant — classic **full mediation**.

---

## Figure 3 — Path diagram

```{r path-diagram, fig.width=10, fig.height=6}
a_r  <- round(a,   3)
b_r  <- round(b,   3)
c_r  <- round(c,   3)
cp_r <- round(c_p, 3)
ab_r <- round(a * b, 3)

ggplot() +
  xlim(0, 10) + ylim(0, 6) +
  coord_fixed() +

  # Boxes
  annotate("rect", xmin = 0.3, xmax = 2.7, ymin = 2.8, ymax = 4.2,
           fill = "#deebf7", color = "#08519c", linewidth = 0.8) +
  annotate("rect", xmin = 3.9, xmax = 6.5, ymin = 4.2, ymax = 5.7,
           fill = "#feedde", color = "#d94801", linewidth = 0.8) +
  annotate("rect", xmin = 7.3, xmax = 9.7, ymin = 2.8, ymax = 4.2,
           fill = "#e5f5e0", color = "#238b45", linewidth = 0.8) +

  # Labels
  annotate("text", x = 1.5, y = 3.7,  label = "Bill length",    fontface = "bold", size = 4.2) +
  annotate("text", x = 1.5, y = 3.2,  label = "X",              color = "grey40",  size = 3.8) +
  annotate("text", x = 5.2, y = 5.15, label = "Flipper length", fontface = "bold", size = 4.2) +
  annotate("text", x = 5.2, y = 4.62, label = "M (mediator)",   color = "grey40",  size = 3.8) +
  annotate("text", x = 8.5, y = 3.7,  label = "Body mass",      fontface = "bold", size = 4.2) +
  annotate("text", x = 8.5, y = 3.2,  label = "Y",              color = "grey40",  size = 3.8) +

  # Arrow a: X → M
  annotate("segment", x = 2.7, xend = 3.85, y = 3.9, yend = 4.6,
           arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
           color = "#d94801", linewidth = 1.1) +
  annotate("text", x = 3.1, y = 4.45, label = paste0("a = ", a_r),
           color = "#d94801", fontface = "bold", size = 4, hjust = 1) +

  # Arrow b: M → Y
  annotate("segment", x = 6.55, xend = 7.25, y = 4.6, yend = 3.9,
           arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
           color = "#d94801", linewidth = 1.1) +
  annotate("text", x = 7.1, y = 4.45, label = paste0("b = ", b_r),
           color = "#d94801", fontface = "bold", size = 4, hjust = 0) +

  # Arrow c': X → Y (direct)
  annotate("segment", x = 2.75, xend = 7.25, y = 3.1, yend = 3.1,
           arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
           color = "#238b45", linewidth = 1.1) +
  annotate("text", x = 5.0, y = 2.75,
           label = paste0("c' = ", cp_r, "  (direct effect, n.s.)"),
           color = "#238b45", fontface = "bold", size = 3.8) +

  # Summary text
  annotate("text", x = 5.0, y = 1.9,
           label = paste0("Indirect effect (a \u00d7 b) = ", a_r, " \u00d7 ", b_r, " = ", ab_r),
           color = "#d94801", fontface = "bold", size = 4.2) +
  annotate("text", x = 5.0, y = 1.35,
           label = paste0("Total effect (c) = ", c_r,
                          "   |   Proportion mediated = ",
                          round(100 * (a * b) / c, 1), "%"),
           color = "grey30", size = 3.8) +

  labs(
    title    = "Mediation path diagram with coefficients",
    subtitle = "93% of bill length's effect on body mass is carried through flipper length",
    caption  = "Data: palmerpenguins | Standardised variables"
  ) +
  theme_void(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle   = element_text(color = "grey40", size = 11, margin = margin(b = 4)),
    plot.caption    = element_text(color = "grey50", size = 9,  margin = margin(t = 8)),
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin     = margin(15, 15, 15, 15)
  )
```

---

## Bootstrap significance test

The {mediation} package tests the indirect effect (a × b) via bootstrapping — no
distributional assumptions required.

```{r bootstrap, cache=TRUE}
set.seed(42)
med_out <- mediate(
  model.m  = model_a,
  model.y  = model_bc,
  treat    = "bill_length_mm",
  mediator = "flipper_length_mm",
  boot     = TRUE,
  sims     = 1000
)
summary(med_out)
```

> **ACME** (Average Causal Mediation Effect) = the indirect effect (a × b): how much of
> bill length's effect on body mass travels *through* flipper length.  
> **ADE** (Average Direct Effect) = the direct effect (c′).  
> **Total Effect** = ACME + ADE = path c.

The bootstrap 95% CI for ACME excludes zero, confirming that the mediated path is
statistically reliable.

---

## Figure 4 — Partial regression plots

Partial regression plots make the story visual. Each plot shows what's left of a predictor
— and of body mass — *after* removing the other predictor's linear influence.

```{r partial-regression, fig.width=11, fig.height=5}
# Residualise bill and body mass after removing flipper's effect
resid_bill_on_flipper <- residuals(lm(bill_length_mm   ~ flipper_length_mm, data = penguins_s))
resid_body_on_flipper <- residuals(lm(body_mass_g      ~ flipper_length_mm, data = penguins_s))

# Residualise flipper and body mass after removing bill's effect
resid_flip_on_bill    <- residuals(lm(flipper_length_mm ~ bill_length_mm, data = penguins_s))
resid_body_on_bill    <- residuals(lm(body_mass_g       ~ bill_length_mm, data = penguins_s))

partial_df <- tibble(
  resid_bill = resid_bill_on_flipper,
  resid_flip = resid_flip_on_bill,
  resid_bm_1 = resid_body_on_flipper,
  resid_bm_2 = resid_body_on_bill
)

slope_bill <- round(coef(lm(resid_bm_1 ~ resid_bill, data = partial_df))["resid_bill"], 3)
slope_flip <- round(coef(lm(resid_bm_2 ~ resid_flip, data = partial_df))["resid_flip"], 3)

p_bill <- ggplot(partial_df, aes(x = resid_bill, y = resid_bm_1)) +
  geom_point(alpha = 0.4, color = "#cb181d", size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, color = "#99000d", linewidth = 0.9) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.5,
           label = paste0("Partial slope = ", slope_bill, " (n.s.)"),
           color = "#99000d", fontface = "bold", size = 4) +
  labs(
    x        = "Bill length | Flipper length (residuals)",
    y        = "Body mass | Flipper length (residuals)",
    title    = "Bill length \u2192 Body mass",
    subtitle = "After removing flipper length from both variables"
  ) +
  theme_blog

p_flip <- ggplot(partial_df, aes(x = resid_flip, y = resid_bm_2)) +
  geom_point(alpha = 0.4, color = "#238b45", size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, color = "#006d2c", linewidth = 0.9) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.5,
           label = paste0("Partial slope = ", slope_flip, " ***"),
           color = "#006d2c", fontface = "bold", size = 4) +
  labs(
    x        = "Flipper length | Bill length (residuals)",
    y        = "Body mass | Bill length (residuals)",
    title    = "Flipper length \u2192 Body mass",
    subtitle = "After removing bill length from both variables"
  ) +
  theme_blog

p_bill + p_flip +
  plot_annotation(
    title    = "Partial regression plots",
    subtitle = "Left: bill length has no unique effect on body mass once flipper is accounted for\nRight: flipper length retains a strong unique effect on body mass",
    caption  = "Data: palmerpenguins | Standardised variables",
    theme    = theme(
      plot.title      = element_text(face = "bold", size = 14),
      plot.subtitle   = element_text(color = "grey40", size = 11),
      plot.background = element_rect(fill = "white", color = NA)
    )
  )
```

The left panel is essentially a scatter of noise — confirming that bill length has no
*unique* variance to offer once flipper length has been accounted for. The right panel
shows a strong, tight relationship — flipper length is doing the real work.

---

## Summary

| | Simple regression | Multiple regression | Mediation |
|---|---|---|---|
| **Bill length β** | 0.595 ✓ | 0.041 ✗ | c′ = 0.041 (direct, n.s.) |
| **Flipper length β** | — | 0.844 ✓ | b = 0.844 |
| **Indirect effect** | — | — | a × b = 0.554 ✓ |
| **Proportion mediated** | — | — | 93.1% |

Multiple regression tells you the coefficient collapses; mediation analysis tells you *the
effect didn't disappear — it just got rerouted*. Bill length predicts flipper length, which
in turn drives body mass. The indirect path accounts for 93% of the total effect, and the
bootstrap confidence interval confirms it is statistically significant.

In plain language: bigger-billed penguins tend to have longer flippers, and longer flippers
predict heavier bodies. Once you know a penguin's flipper length, knowing the bill length
adds very little extra predictive value for body mass.

---

*Analysis by Rahul Venugopal | Data: {palmerpenguins} | All variables standardised*
